---
title: "Surplus Production Demo"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
```{r functions}
library(shiny)
library(ggplot2)
library(gridExtra)
func1<-function(params) {
    df1<-swo
    numyears<-dim(df1)[1]
    B0.K<-1
    df1$Biomass[1]<-params["K"]*B0.K
    df1$Est.CPUE[1]<-df1$Biomass[1]*params["q"]
    for(y in 2:numyears) {
     df1$Biomass[y]<-max(c(df1$Biomass[y-1]+params["r"]*df1$Biomass[y-1]*(1-df1$Biomass[y-1]/params["K"])-df1$Catch[y-1],0.0001))
     df1$Est.CPUE[y]<-df1$Biomass[y]*params["q"]
     df1$SS[y]<-(df1$CPUE[y]-df1$Est.CPUE[y])^2
    }
    sum(df1$SS,na.rm=TRUE)
    }
func2<-function(df1,params) {
    numyears<-dim(df1)[1]
    B0.K<-1
    df1$Biomass[1]<-params["K"]*B0.K
    df1$Est.CPUE[1]<-df1$Biomass[1]*params["q"]
    for(y in 2:numyears) {
     df1$Biomass[y]<-max(c(df1$Biomass[y-1]+params["r"]*df1$Biomass[y-1]*
         (1-df1$Biomass[y-1]/params["K"])-df1$Catch[y-1],0.0001))
     df1$Est.CPUE[y]<-df1$Biomass[y]*params["q"]
     df1$SS[y]<-(df1$CPUE[y]-df1$Est.CPUE[y])^2
      }
     df1 
    }
func3<-function(df1,params,HRtype,HR,Quota) {
    numyears<-dim(df1)[1]
    B0.K<-1
    df1$Biomass[1]<-params["K"]*B0.K
    for(y in 2:numyears) {
     df1$Biomass[y]<-max(c(df1$Biomass[y-1]+params["r"]*df1$Biomass[y-1]*(1-df1$Biomass[y-1]/params["K"])-df1$Catch[y-1],0.0001))
    }
    df1<-rbind(df1,data.frame(Year=rep(0,projnum),Catch=rep(0,projnum),
      CPUE=rep(NA,projnum),Biomass=rep(0,projnum)))
    for(y in (numyears+1):(numyears+projnum)) {
     df1$Year[y]<-df1$Year[y-1]+1
     df1$Catch[y]<-ifelse(HRtype=="Fraction caught",HR*df1$Biomass[y-1],Quota)
     df1$Biomass[y]<-max(c(df1$Biomass[y-1]+params["r"]*df1$Biomass[y-1]*
         (1-df1$Biomass[y-1]/params["K"])-df1$Catch[y-1],0))
     }
     df1 
    }

```

This demo runs a very simple populations dynamics model, the Schaefer surplus production model, which has a yield curve where the maximum sustainable yield (MSY) occurs at half of the carrying capacity (K).

```{r}
df1<-data.frame(Biomass=0:1000)
df1$Yield<-0.5*df1$Biomass*(1-df1$Biomass/1000)
ggplot(df1,aes(x=Biomass,y=Yield))+geom_line(col="darkgreen",linewidth=2)+
  ylab("Surplus production = sustainable yield")+
  xlab("Increasing Biomass or Decreasing Fishing Mortality -->")+
  theme_bw(base_size=12)
```

The goals of this demo are to: 

1. Understand reference points related to Maximum Sustainable Yield,

2. Compare fixed catch vs fixed harvest rate strategies, and 

3. Develop intuition for how fish populations respond to exploitation. 

## Production model with a constant catch quota

The model below is set with K=1000, and r= 0.5, which means that the biomass that can support Maximum Sustainable Yield is:
$$B_{MSY}=K/2=500$$
the harvest fraction is:
$$H_{MSY}=r/2=0.25$$
and maximum sustainable yield is:
$$MSY=rK/4=125$$

The sliders let you adjust the starting biomass relative to K and the annual catch quota.

Try the following scenarios. What happens to the biomass over time?

**1. Sustainable catch levels for a new fishery** Set the starting biomass equal to the carrying capacity ($B_0/K=1$), as would be expected in a population that had never been fished. Now try several different levels of annual quota. 

a. Try an annual quota higher than the MSY.

b. Try an annual quota lower than the MSY?

**2. Sustainable catch levels for a depleted fishery**  Now set the starting biomass $B_0/K$ at 0.1, a severely overfished population. What is the highest quota that is sustainable?    

```{r constant catch}
inputPanel(
      sliderInput(inputId = "B0.K",
                  label = "Starting Biomass/K:",
                  min = 0,
                  max = 1,
                  value = 0.1),
      sliderInput(inputId = "Quota",
                  label = "Annual Catch Quota:",
                  min = 0,
                  max = 200,
                  value = 0)
    )
renderPlot({
   numyears<-50
   K<-1000
   r<-0.5
   MSY<-r*K/4
   Bmsy<-K/2
   Hmsy<-r/2
   df1<-data.frame(Year=1:numyears)
   df1$Biomass[1]<-K*input$B0.K
   df1$Catch[1]<-input$Quota
   for(y in 2:numyears) {
    df1$Biomass[y]<-max(c(df1$Biomass[y-1]+r*df1$Biomass[y-1]*(1-df1$Biomass[y-1]/K)-df1$Catch[y-1],0))
    df1$Catch[y]<-min(c(df1$Biomass[y],input$Quota))
   }
   ggplot(df1,aes(x=Year))+geom_line(aes(y=Biomass),col="blue",linewidth=2)+
    geom_line(aes(y=Catch),col="red",linewidth=2)+theme_bw(base_size=12)+
     ylab("Catch and Biomass")+
     ggtitle("With Constant Catch Quota, r=0.5,K=1000")+
     geom_text(x=numyears-10,y=K*0.9,label="Biomass",col="blue")+
     geom_text(x=numyears-10,y=K*0.8,label="Catch",col="red")+
     ylim(c(0,K))
    })
```

## Production model with a fixed harvest fraction

The following is the same production model, but instead of having a total allowable catch quota, you can adjust the fraction of the population caught every year (i.e. Catch/Biomass), often called the harvest rate (H). 

**3. Starting from Bmsy** Set $B_0/K$  equal to 0.5 and set the catch rate equal to any number between zero and 1. What catch rate prevents the population from declining?

**4. Starting with a depleted population**  Set $B_0/K$ equal to 0.1. What catch rate prevents the population from declining? 

```{r harvest rate}
inputPanel(
       sliderInput(inputId = "B0.K2",
                  label = "Starting Biomass/K:",
                  min = 0,
                  max = 1,
                  value = 0.1),
      sliderInput(inputId = "F",
                  label = "Fraction caught:",
                  min = 0,
                  max = 0.5,
                  value = 0)
)

renderPlot({
   numyears<-50
   K<-1000
   r<-0.5
   MSY<-r*K/4
   Bmsy<-K/2
   Hmsy<-r/2
   df1<-data.frame(Year=1:numyears)
   df1$Biomass[1]<-K*input$B0.K2
   df1$Catch[1]<-input$F*df1$Biomass[1]
   for(y in 2:numyears) {
    df1$Biomass[y]<-max(c(df1$Biomass[y-1]+r*df1$Biomass[y-1]*(1-df1$Biomass[y-1]/K)-df1$Catch[y-1],0))
    df1$Catch[y]<-min(c(df1$Biomass[y],input$F*df1$Biomass[y]))
   }
   #Make plot
   ggplot(df1,aes(x=Year))+geom_line(aes(y=Biomass),col="blue",linewidth=2)+
    geom_line(aes(y=Catch),col="red",linewidth=2)+theme_bw(base_size=12)+
     ylab("Catch and Biomass")+
     ggtitle("With a constant harvest rate strategy")+
     theme_bw(base_size=12)+
     geom_text(x=numyears-10,y=K*0.9,label="Biomass",col="blue")+
     geom_text(x=numyears-10,y=K*0.8,label="Catch",col="red")+
     ylim(c(0,K))
    })

```


## Fitting the model to data

To use a production model for stock assessment, we use statistical methods to estimate the parameters r and K so that we can calculate the biomass in each year from the surplus production minus the catch. We also estimate catchability (q), which is the constant of proportionality between the catch per unit effort (CPUE) and the biomass [^1]. For example, We can estimate the parameters by minimizing the sum of squared (SS) difference between the observed and predicted catch per unit of effort (CPUE). To see how this works in practice, we will fit the data for North Atlantic swordfish from ICCAT(1999).

```{r, plot swordfish,warning=FALSE}
swo<-read.csv("Swordfish.csv")
swo1<-ggplot(swo,aes(x=Year,y=CPUE))+
      geom_point(color="darkred")+theme_bw(base_size=12)
swo2<-ggplot(swo,aes(x=Year,y=Catch))+
      geom_col(fill="darkred")+theme_bw(base_size=12)
grid.arrange(swo1,swo2,ncol=2)
```

**6. Fit the model to the swordfish data ** First use the sliders to explore what values of r, K and catchability (q, the constant of proportionality between the index and the biomass) best fit the data. After you have tried several values, select "Solve" to let R solve for the values that minimize the sum of squares (SS). Does the fit look reasonable?

```{r solve}
swo<-read.csv("Swordfish.csv")
inputPanel(
      # Input: Slider for r ----
      sliderInput(inputId = "r",
                  label = "Intrinsic growth rate(r):",
                  min = 0,
                  max = 1,
                  value = 0.8),
      sliderInput(inputId = "K",
                  label = "Carrying Capacity(K):",
                  min = 10,
                  max = 150000,
                  value = 70000),
      sliderInput(inputId = "q",
                  label = "Catchability (q):",
                  min = 0.00001,
                  max = 0.01,
                  value = 0.005),
      selectInput("SolveOpt","Solve?",choices=c("Use slider","Solve"),
        selected = "Use slider")
    )
plot1<-reactive({
     df2<-func2(swo,c(q=input$q,K=input$K,r=input$r))
     ggplot(df2,aes(x=Year))+geom_line(aes(y=Biomass*input$q),col="blue",linewidth=2)+
      geom_point(aes(x=Year,y=CPUE))+
      geom_text(aes(x=1985,y=1200,label=paste("SS=",round(sum(SS,na.rm=TRUE),1))))+
      ylab("CPUE (point) and abundance trend (line)")+theme_bw(base_size=12)
})
plot2<-reactive({
     fit1<-nlminb(start=c(q=0.001,K=70000,r=0.8),objective=func1,lower=0)
     df2<-func2(swo,fit1$par)
     ggplot(df2,aes(x=Year))+geom_line(aes(y=Biomass*fit1$par["q"]),col="blue",linewidth=2)+
      geom_point(aes(x=Year,y=CPUE))+
      geom_text(aes(x=1985,y=1000,label=
         paste0("SS=",round(sum(fit1$objective,na.rm=TRUE),1),
       "\n r=",round(fit1$par["r"],2),"\n K=",round(fit1$par["K"],1),
       "\n q=",round(fit1$par["q"],4))),size=8)+
       ylab("CPUE and abundance trend")+theme_bw(base_size=12)
    })
renderPlot({
      switch(input$SolveOpt, 
        "Use slider"=plot1(), 
        "Solve"=plot2()
      )
  })

```

## Projections from the production model

**7. Projections** In the following, you can select a harvest control rule and project the population forward 10 years with our estimated values of r and K, with either a fixed catch quota or a catch fraction. 

a. What catch quota or catch fraction will rebuild the population to Bmsy in 10 years? 

b. Are catch quotas or catch fractions better?

```{r}
swo<-read.csv("Swordfish.csv")
projnum<-10
inputPanel(
  selectInput("RuleType","Quota or harvest rate?",
        choices=c("Quota","Fraction caught"),
        selected = "Quota"),
    sliderInput(inputId = "TAC",
                  label = "Constant catch quota:",
                  min = 0,
                  max = 10000,
                  value = 0),
      sliderInput(inputId = "HR",
                  label = "Fraction caught (Harvest rate):",
                  min = 0,
                  max = 1,
                  value = 0)
    )

renderPlot({
     fit1<-nlminb(start=c(q=0.001,K=70000,r=0.8),objective=func1,lower=0)
     Bmsy<-fit1$par["K"]/2
     Hmsy<-fit1$par["r"]/2
     df3<-func3(swo,fit1$par,input$RuleType,input$HR,input$TAC)
     df3$H<-df3$Catch/df3$Biomass
     ggplot(df3,aes(x=Year))+geom_line(aes(y=Biomass/Bmsy),col="blue",linewidth=2)+
      geom_line(aes(x=Year,y=H/Hmsy),linewidth=2,col="red")+
      theme_bw(base_size=12)+
      ylab("B/Bmsy and H/Hmsy")+
      geom_hline(yintercept=1,lty=2)+
      geom_vline(xintercept=1999.5)+
      geom_text(x=2000,y=1.9,label="Begin \nprojections",hjust=0,size=4)+
      geom_text(x=1950,y=1.5,label="Biomass/Bmsy",col="blue",hjust=0)+
      geom_text(x=1950,y=1.3,label="H/Hmsy",col="red",hjust=0)
    })
```

## Conclusion

This demo shows the behavior of a very simple population dynamics model. It shows that catches have to be much lower than the maximum sustainable yield to rebuild an overfished population, and that the constant harvest rates rebuild faster than constant catch quotas because they respond to changes in biomass. In practice, the population dynamics of any fish species would include more variability and uncertainty than we allowed in this model, as, for example, environmental conditions can cause good and bad recruitment years, variability in growth rates, and fish movement. Also, managers would not be able to control catch perfectly. Thus, in practice, fisheries are generally not managed for MSY, but rather for more precautionary reference points. Also, most fisheries are assessed with age structured modes to account for size-selectivity, time lags for slow growing species with a late age at reproduction, and other elements of the population dynamics. 

For questions and feedback on this demo, please contact ebabcock@miami.edu

[^1]: For most methods, we also need to estimate the measurement error in CPUE. we will assume that the measured CPUE is normally distributed around predicted CPUE. We can either estimate the biomass in the first year, or assume it is equal to K, as in this case. 
